Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-23 ind_files/04_field_eqns.tex
Minor rephrasing
Add a reference for martingale problem

The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes I study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\citet{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future.

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a very big ask~\citep{dawson_stochastic_1975,walsh_introduction_1986,etheridge_mathematical_2011,rogers_demographic_2012}.

>>>>

The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes I study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\citet{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future.

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a very big ask~\citep{dawson_stochastic_1975,walsh_introduction_1986,dawson_stochastic_2000}.

### 2023-03-23 ind_files/05_unification_discussion.tex
Add a reference for martingale problem

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful and has been extensively exploited in the literature~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because some problems are much more easily attacked using tools from probability theory that operate on It\^o processes, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example), fast-mode (`adiabatic') elimination/slow manifold approximation~\citep{constable_stochastic_2013,parsons_dimension_2017}, and WKB style expansions to detect extreme events~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and often come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators/semi-group theory~\citep{ethier_markov_1986,etheridge_mathematical_2011,finkelshtein_semigroup_2012}, and the martingale problem~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}. Rigorously establishing and systematically exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and the sort of general SPDEs studied by~\citet{week_white_2021} for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{The physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in completely `classical' systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

>>>>

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful and has been extensively exploited in the literature~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because some problems are much more easily attacked using tools from probability theory that operate on It\^o processes, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example), fast-mode (`adiabatic') elimination/slow manifold approximation~\citep{constable_stochastic_2013,parsons_dimension_2017}, and WKB style expansions to detect extreme events~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and often come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators/semi-group theory~\citep{ethier_markov_1986,finkelshtein_semigroup_2012}, and the martingale problem~\citep{ethier_markov_1986,dawson_stochastic_2000, champagnat_unifying_2006}. Rigorously establishing and systematically exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and the sort of general SPDEs studied by~\citet{week_white_2021} for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{The physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in completely `classical' systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.
